Generalized Algebraic Data Types in Haskell

### 2013-08-16 gadt.tex
Added a reference to the expression problem and a reference to the paper "Data types a la carte" where a solution is described.
Fixed typos and errors in the article

\begin{Verbatim}
evaluate e = case e of
    AddInt e1 e2 -> case (evaluate e1, evaluate e2) of
        (Just (Left i1), Just (Left i2)) -> Just $ Left $ i1 + i2
        _ -> error "AddInt takes two integers"
\end{Verbatim}

@@

These invariants guarantee that tree is balanced. Indeed, the longest path from the root node (containing alternating red-black nodes) can only be twice as long as the shortest path (containing only red nodes). Thus basic operations (such as insertion and search) take $O(\log n)$ time in the worst case.

@@

\begin{Verbatim}
type Tree a = forall bh. Node a bh
\end{Verbatim}

@@

Another important aspect is that GADTs, just like all data types in Haskell, are closed: once they are declared, they can not be extended in the rest of the program. In contrast, type class instances are open. For example, we could split \verb|Expr e| type class into two separate type classes (this way we improve modularity):

\EZY{You should probably cite Data Types a la Carte here}

>>>>

\begin{Verbatim}
evaluate e = case e of
    AddInt e1 e2 -> case (evaluate e1, evaluate e2) of
        (Just (Left i1), Just (Left i2)) -> Just $ Left $ i1 + i2
        _ -> fail "AddInt takes two integers"
\end{Verbatim}

@@

These invariants guarantee that tree is balanced. Indeed, the longest path from the root node (containing alternating red-black nodes) can only be twice as long as the shortest path (containing only black nodes). Thus basic operations (such as insertion and search) take $O(\log n)$ time in the worst case.

@@

\begin{Verbatim}
data Tree a = forall bh. Root (Node a bh)
\end{Verbatim}

@@

Another important aspect is that GADTs, just like all data types in Haskell, are closed: once they are declared, they can not be extended in the rest of the program. This issue is called the \italic{expression problem}~\cite{expression_problem}. Wouter Swierstra describes a technique to solve this problem in the article ``Data types \`{a} la carte''~\cite{a_la_carte}.

In contrast, type class instances are open. For example, we could split \verb|Expr e| type class into two separate type classes (this way we improve modularity):
